FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2023-08-11 paper.tex
Major rework of section 2

%% TODO use the GPU-FPX paper as a guide for the background. Give a little more detail than that, but not much!

% TODO: talk about how floating point != real numbers → errors appear and build up
% (see other tutorials; Art of Computer Programming, Ganesh, slides in Google Drive from guy in England)
% 2 paragraphs or so: floating point primer

\Fp{} numbers are a way of representing real numbers in a computer.
However, since we limit our precision to a fixed, finite set of bits, \fp{} values are \emph{inherently approximate} and necessarily incur some error during computation.
This is not bad; it is a trade-off: we get fast and compact representations that are good enough for most use cases.
However, the downside is that \fp{} numbers \emph{do not} behave like real numbers.

% FIXME: more resources!
There are many excellent resources\cite{knuthArtComputerProgramming1997,torontoPracticallyAccurateFloatingPoint2014} on understanding the theory behind \fp{};
we will explain the most essential aspects here briefly.

A \fp value approximates real number $r \in \mathbb{R}$ by expressing $r$ in scientific notation $r = s \times f \times b^{e-q}$,
for some base $b$,
an integer exponent $e \in \mathbb{N}$,
a fixed value called an \emph{excess} $q$,\footnote{The purpose of the excess $q$ is so that with value of $e > q$ we can represent positive exponents, and $e < q$ we can represent negative exponents.}
a \emph{sign} $s \in \{-1, 1\}$,
and a \emph{fraction} $f$ such that $0 \leq f < b$.\footnote{Another name for the fraction part commonly seen in the literature is ``mantissa'' or ``significand''.
  You might occasionally also see ``characteristic'' referring to the exponent.
  We will stick to ``fraction'' and ``exponent'' however.}

We pack $s$, $e$, and $f$ into bit fields.
This gives us a compact representation for both very small and very large values: because the exponent is variable, we can represent numbers as big as Avogadro's number or as small as Planck's constant with the same number of bits.
However, since we have only a finite number of bits $p$, we have to round $f$ and the range of $e$ is limited.
Moreover, there is some error between the real number $r$ and the value we can express with a float.
This is why $0.1 + 0.2 = 0.30000000000000004$---$0.3$ is not precisely representable.

To make matters worse, things get tricky when we do math.
Adding e.g. Avogadro's number and Planck's constant together is beyond 16-, 32-, and 64-bit floats to represent.
In fact, whenever the exponents differ by more than $p+2$, the result of the addition will just be the same as the larger number.
In the case of Avogadro's number, adding any value less than $FIXME$ will just result in Avogadro's number again.

The gradual accumulation of these errors and other representation issues leads to surprising results!
% FIXME: more examples of a NaN popping up out of nowhere, e.g. with bad addition

In our work we focused on two of the most common problematic values in the \fp{} space: \NaN{} and \Inf{}.
We leave the possibility of tracking other kinds of troublesome values (e.g. subnormal numbers) open, but for our work we restrict ourselves to just these values.

\subsection{Where do \NaN{}s come from?}

IEEE 754~\cite{IEEEStandardBinary1985} lists the following operations (among a few others) where \NaN{}s result:

@@

When one of these events happens, we call it a \NaN{} \emph{generation}.

Most mathematical operations preserve \NaN{}s.
We call these instances \NaN{} \emph{propagation}, as the \NaN{} is propagated from the source in the arguments to the result.

There are, however, some instances where the IEEE standard requires that \NaN{}s to \emph{not} propagate from arguments to the result; we call these \NaN{} \emph{kills}.
Sometimes this can be good---an \texttt{isnan} check should clearly return a boolean and not a \NaN{}---but kills are often harmful in two ways:

\begin{enumerate}
  \item Users no longer see that there was some errant operation: \NaN{}s are an opportunity to closely examine data for problems and code for unsoundness.
  \item \NaN{} kills can influence the control of a program leading to incorrect results, as we saw in~\cref{s:to-kill-a-fp}.
\end{enumerate}

Here are cases where, according to the IEEE standard, a \NaN{} can be killed:

\begin{align*}
x \; \texttt{cmp} \; y &\rightarrow false &\text{where \texttt{cmp} is \texttt{<}, \texttt{>}, \texttt{<=}, etc.} \\
1^{\text{\texttt{\NaN{}}}} &\rightarrow 1 \\
\text{\texttt{\NaN{}}}^{0} &\rightarrow 1
\end{align*}

FloatTracker monitors these and all other operations for interesting events.

\subsubsection{TODO good quote, try to work this in}

From IEEE 754. Section 6.2 here: \url{http://www.dsc.ufcg.edu.br/~cnum/modulos/Modulo2/IEEE754\_2008.pdf}:

\begin{quote}
  Quiet NaNs should, by means left to the implementer’s discretion, afford
  retrospective diagnostic information inherited from invalid or unavailable
  data and results.
\end{quote}

Found via \url{https://anniecherkaev.com/the-secret-life-of-nan}

TODO what other tools are there for retrospective diagnosis?

\subsection{Infinity}

%%ben: shouldn't this come before NaN b/c infs lead to NaNs?

There are two infinities, $+\infty$ and $-\infty$, but they behave the same for our purposes, so we will refer to them just as ``\Inf{}''.
Like \NaN{}s, \Inf{} can crop up when bad math happens: $\frac{x}{0} \rightarrow \infty$ when $x \ne 0$.
\Inf{} can also arise when numbers get too big or too small for the choice of representation.

@@

@@

The IEEE 754 specification for \fp{} numbers governs how \NaN{}s are handled as arguments to arithmetic functions~\cite{IEEEStandardBinary1985}.
Most of the time, when any or all of the arguments are \NaN{}, the result is also a \NaN{}.
This is good and desirable, because it means that an arithmetic error---even if buried deep in an expression---will still be observable, allowing a user to find and debug it.

However, there are a few places where \NaN{}s flow into an expression, but not out.
We call these instances ``\NaN{}-kills'', and these can lead to subtle logic bugs.
Consider the following example:

>>>>

\Fp{} provides a finite approximation of the real numbers in a format that is efficient for calculation on a computer.
IEEE 754~\cite{IEEEStandardBinary1985} standardizes \fp{} representation and operations---these days, most machines and languages follow the IEEE 754 standard, and we make no effort to generalize our work beyond what IEEE 754 specifies.

\Fp{} is essentially scientific notation---\fp{} numbers are represented by packing a sign bit, an exponent, and an fraction part (also called the ``significand'' or ``mantissa'') into contiguous bit fields.
The advantages of this representation are that allows for compact representation of both very small and very large numbers in a narrow range of bits---a 64-bit float can represent numbers as big as $2^{1024}$ and as small as $2^{-1074}$.
This range is sufficient to represent both Planck's constant as well as Avogadro's constant with reasonable accuracy, making 
The flip side is that not every real number within that wide range is representable as a \fp{} value---there is almost always some gap between the real number that we wish to represent and the \fp{} number that can represent.
Whenever a value we have doesn't have a precise representation, we must \emph{round} that real value to the nearest \fp{} value.
We call this gap error, and error, along with the finite nature of \fp{} values, means that we can run into some funny situations.

For example, if two numbers' exponents differ by too large an amount, adding or subtracting the smaller from the bigger can result in just the bigger number.
For example, $6.022 \times 10^{23} + 6.626 \times 10^{-34}$ (Avogadro's constant plus Planck's constant) will leave with with just Avogadro's constant---there is no way to pack all the significant figures from true value that differs minutely from Avogadro's number into a 64-bit \fp{} number.
This is a fairly extreme example; even when the numbers are closer in magnitude, most operations on \fp{} numbers induce loss of accuracy.

\subsection{Exceptions and exceptional values}

IEEE 754 defines \emph{\fp{} exceptions}, which are events that occur when something goes wrong in a calculation.
William Kahan says,

\begin{quote}
An arithmetic exception arises when [there is] no result that would be acceptable universally.
\end{quote}

For example, dividing by zero is an exception, as is exceeding your \fp{} format's range.
These are exceptions because it is impossible to say what the right answer \emph{should} have been: perhaps that zero was meant to be a value that got really \emph{close} to zero, and suddenly rounded towards zero before division.
Or maybe that value that got really big \emph{does} need to get that big, but you need to start using a larger format for your \fp{} values.
Whatever, the case, there isn't a universally acceptable answer, so a \fp{} exception gets raised.

IEEE 754 doesn't always obligate the programmer to handle these exceptions, however, so we end up with \emph{exceptional values}, which are the results of exception-raising operations.
The two values we work with in this paper are \Inf{} and \NaN{}, which result from overflow and nonsensical operations respectively.\footnote{The third kind of exceptional value is called a \emph{subnormal} value. Please see~\cite{torontoPracticallyAccurateFloatingPoint2014} for a good introduction to subnormal values, as well as other details of \fp{} representation.}

IEEE 754 lists the following operations (among a few others) where \NaN{}s result:

@@

There are actually two infinities, $+\infty$ and $-\infty$, but they behave the same for our purposes, so we will refer to them just as ``\Inf{}''.
Like \NaN{}s, \Inf{} can crop up when bad math happens: $\frac{x}{0} \rightarrow \infty$ when $x \ne 0$.
\Inf{} can also arise when numbers get too big or too small for the choice of representation.

There are many excellent resources\cite{knuthArtComputerProgramming1997,torontoPracticallyAccurateFloatingPoint2014} on understanding the theory behind \fp{} as well as how error is measured and accumulates.
Please see those works for more details.
What we have described above should suffice for the remainder of this paper.

@@

Exceptional values have a life cycle: they are born, live, and die, as seen in \cref{f:gpk}.

@@

When a \NaN{} or an \Inf{} comes into being as the result of a bad math operation, we call this a generation, or \emph{gen} for short.

The IEEE 754 specification for \fp{} numbers governs how \NaN{}s are handled as arguments to arithmetic functions~\cite{IEEEStandardBinary1985}.
Most of the time, when any or all of the arguments are \NaN{}, the result is also a \NaN{}.
We call this a propagation, or \emph{prop} for short.

This is good and desirable, because it means that an arithmetic error---even if buried deep in an expression---will still be observable, allowing a user to find and debug it.

However, there are a few places where exceptional values flow into an expression, but not out.
We call these instances \emph{kills}, and these can lead to subtle logic bugs.
For example, \emph{any} comparison ($<$, $>$, $=$, $\neq$, etc.) with a \NaN{} returns \texttt{false}.

Consider the following two functions to find the maximum value in a list:
